---
title: "Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl 
fontsize: 11pt
geometry: margin=1.0in
---


```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
source('../code/functions.R')

loadLibs(c("tidyverse", "knitr", "rmarkdown"))
```


```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

```


\vspace{35mm}

Running title: SCFAs and Colorectal Cancer

\vspace{35mm}


Marc A Sze${^1}$, Nicholas A Lesniak${^1}$, Mack T Ruffin IV${^2}$, Patrick D. Schloss${^1}$${^\dagger}$

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI 48109

$2$ Department of Family Medicine and Community Medicine, Penn State Hershey Medical Center, Hershey, PA


\newpage
\linenumbers


## Abstract


\newpage

## Introduction


\newpage


## Results 

\newpage

## Discussion

\newpage

## Conclusions

\newpage

## Materials and Methods


\newpage

## Acknowledgements

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Kwi Kim and Thomas M Schmidt for their help in running the short-chian fatty acid analysis on the High-Performace Liquid Chromatography machine at the University of Michigan. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.   

\newpage

## References

<div id="refs"></div>

\newpage

Insert figure legends with the first sentence in bold, for example:

**Figure 1. Number of OTUs sampled among bacterial and archaeal 16S rRNA gene sequences for different OTU definitions and level of sequencing effort.** Rarefaction curves for different OTU definitions of Bacteria (A) and Archaea (B). Rarefaction curves for the coarse environments in Table 1 for Bacteria (C) and Archaea (D).


\newpage
